Building and Checking Suffix Array Using Induced Sorting Method

### 2017-06-27 new-draft/draft.tex
20170628

%The induced sorting~(IS) method has been applied to designing several disk-based suffix array~(SA) construction algorithms. Although these algorithms have better time and I/O complexities than the other alternatives, their current programs commonly suffer from a bottleneck due to the large disk space in need.

The induced sorting~(IS) method has been applied to designing efficient suffix array~(SA) construction algorithms on both internal and external memory models. In this paper, we propose two checking methods that enable any IS suffix sorters to check an SA when it is being built. The first method is capable of checking both suffix and LCP arrays for strings on constant alphabets, while the second method is specific for checking SA only but applicable to strings on both constant and integer alphabets. For performance analysis, we integrate the second checking method into the existing SA construction algorithm DSA-IS, where the design of DSA-IS is adapted by new substring sorting and naming methods for reducing its peak disk use to at most 14 times as input size. From our experimental results, the time, space and I/O consumptions for verification by our second method is negligible to that for construction by the adapted DSA-IS. This convinced that our checkers, together with the state-of-the-art IS suffix sorters, may constitute efficient solutions for building and checking SA at the same time.

@@

The suffix array of any constant or integer string can be built in linear time and RAM space by SA-IS~\cite{Nong11}. The induced-sorting method has been also applied to designing fast external-memory SA construction algorithms eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}. These three disk-based variants have better time complexities than the other alternatives~(e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}), but they all suffer from a bottleneck due to the large disk space in use. For example, when building SA encoded by 40-bit integers, the peak disk use for eSAIS, DSA-IS and SAIS-PQ are $24n$, $20n$ and $15n$, respectively. This situation changed due to the work presented in~\cite{Karkkainen2017}, which proposed a carefully engineered IS suffix sorter fSAIS that takes no more than $8n$ disk memory for any constant or integer string of size $n \le 2^{40}$~\cite{Karkkainen2017}. As reported, fSAIS reuses the substring/suffix sorting and naming methods of DSA-IS and SAIS-PQ, it achieves the remarkable space performance by means of a monotone priority queue for sorting fixed-size tuples and an I/O layer for allocating and deallocating disk space in a fine-grained way. The successful design of fSAIS shows a great potential for sharply decreasing the disk space requirements of the current programs for DSA-IS and SAIS-PQ. In this paper, we present our first attempt to improve DSA-IS by new substring sorting and naming methods. From the experiments, our program for the adapted algorithm DSA-IS+ takes $12n$ disk space in average and runs as fast as eSAIS on various real-world datasets. We will demonstrate that this program can be further improved using the optimization technique for implementing fSAIS.

To avoid computation errors caused by implementation bugs and/or hardware malfunctions, an SA should be checked to ensure its correctness after construction. Currently, the software packages for DC3 and eSAIS provide users a checker based on the method proposed in~\cite{Dementiev2008a}, which performs two disk sorts for $\mathcal{O}(n)$ fixed-size tuples with an integer key. When running in external memory, this checker takes one-third as much time and I/O consumptions as the programs for DSA-IS and eSAIS, it also requires nearly the same disk space as the two SA builders. Against the background, we propose two checking methods that enable any IS suffix sorting algorithm to build and check SA at the same time. In comparison with the previous checking method, they are rather lightweight in terms of time, space and I/O volume. It should be noticed that these two methods are designed for different situations. In brief, the first method can check both suffix and LCP arrays, but it is specific for strings on constant alphabets. The second method can be applied to strings on constant and integer alphabets, but it can check SA only. As will be seen in our experiments, the consumption for checking by our second method is negligible compared with that for building by DSA-IS+.

The rest of this paper is organized as follows. Section~\ref{sec:builder} describes the framework of IS suffix sorting algorithms and the idea of new substring sorting and naming methods, Section~\ref{sec:checkers} the designs of two checking methods, Sections~\ref{sec:experiments} the experimental results, and Section~\ref{sec:conclusion} the concluding remarks.

@@

\subsection{Preliminaries}

Consider a string $x[0,n)$ drawn from a full-ordered alphabet $\Sigma$. The ending character $x[n - 1]$ is supposed to be unique and lexicographically smaller than any others in $x$. For convenience, we denote by ${\sf suf}(i)$ the suffix running from $x[i]$ to the ending character and ${\sf sub}(i,j)$ the substring running from $x[i]$ to $x[j]$, respectively. The following notations are also used in our presentation.

{\em Character/substring/suffix classification.} Characters in $x$ are classified into two categories: L-type and S-type. We say $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] = x[i + 1]$ and $x[i + 1]$ is S-type; otherwise $x[i]$ is L-type. Furthermore, if $x[i]$ and $x[i - 1]$ are respectively S-type/L-type and L-type/S-type, then $x[i]$ is also called S*-type/L*-type. An array $t$ records the type of all the characters in $x$, $t[i]$ is set to $1$ or $0$ if $x[i]$ is S-type or L-type, respectively. Each substring/suffix is of the same type as its heading character. In this paper, we only consider the substrings ending with an S*-type character.

{\em Predecessor and successor.} Given two characters $x[i]$ and $x[i + 1]$, $x[i]$ is the predecessor of $x[i + 1]$ and $x[i + 1]$ is the successor of $x[i]$. Accordingly, we define the predecessor-successor relationship between two suffixes or substrings starting with neighboring characters.

{\em Reduced string.} Split $x$ into multiple substrings such that each substring only contains two S*-type characters at the starting and ending positions. Assign each substring a name and replace them with their names to produce a reduced string $x_1$, where the name indicates the lexical order of the corresponding substring among all.

{\em Suffix array.} The suffix array $sa$ indicates the lexicographical order of all the suffixes in $x$, where $sa[i]$ records the starting position of the $(i + 1)$-th smallest suffix. We also define the SA for $x_1$ by $sa_1$.

@@

	\caption{The general framework for an IS suffix sorting algorithm.}

@@

As shown in Algorithm~\ref{alg:1}, an IS suffix sorting algorithm first performs a reduction phase to produce the reduced string $x_1$. If there exist two equal characters in $x_1$, then it recursively calls the reduction phase with $x_1$ as input; otherwise, all the S*-type suffixes in $x$ are already sorted and it performs an induction phase to produce $sa$ from these sorted suffixes. According to the IS principle, the order of two substrings/suffixes is determined by comparing their heading characters and successors in sequence during the execution of reduction/induction phase. Notice that this key operation takes multiple random accesses to $x$ and $sa$. It can be done very quickly if both $x$ and $sa$ are wholly loaded into RAM; otherwise, each access may take an individual I/O operation to external memory and thus becomes the bottleneck.

To amortize the I/O overhead, eSAIS, DSA-IS and SAIS-PQ use different methods to obtain heading characters in a disk-friendly way. Particularly, DSA-IS first divides $x$ into multiple blocks and calls SA-IS to sort suffixes in each block at the beginning of induction phase, where the heading characters in need are written to external memory in their access order. Then, it organizes the sorted suffixes in a priority queue and visits them sequentially to induce the order of their predecessors following the IS principle. Because the required heading characters was already stored on disks, they can be read into RAM by sequential I/O operations when needed. This technique is also applied to sorting substrings during the reduction phase. Afterward, a naming process is conducted to check equality of lexicographically neighboring substrings for producing the reduced string $x_1$. To do this, DSA-IS maintains a copy for each substring in external memory and arranges them in their sorted order. As can be seen in Section~\ref{sec:experiments}, our program for DSA-IS requires less disk space than that for eSAIS, but it runs slower than the latter due to the large I/O volume for sorting and naming S*-type substrings.

@@

We first describe a new method for sorting S*-type substrings during the reduction phase. Given an integer value $D$, all the S*-type substrings are classified into long and short categories with respect to whether or not they contain more than $D$ characters. Our sorting method executes the following three steps in sequence after splitting $x$ into blocks $\{b_1, b_2, ...\}$.

\begin{enumerate}[S1]
	\item Sort S*-type substrings in each block by calling the reduction phase of SA-IS. Copy the short substrings and store them on the disk in their sorted order. Copy the heading characters needed for sorting long substrings and store them on the disk in their access order.~\label{sort_and_name_method:1}

	\item Sort long S*-type substrings by the same way of DSA-IS. Copy the leftmost $D$ characters of the long substrings and store them on the disk in their sorted order.~\label{sort_and_name_method:2}

	\item Merge short and long S*-type substrings by a multi-way sorter.~\label{sort_and_name_method:3}
\end{enumerate}

Assume the number of blocks is $k$, S\ref{sort_and_name_method:1} copies the sorted short substrings in $b_i$ to an external memory vector $vs_i$ and S\ref{sort_and_name_method:2} copies the sorted long substrings to an external memory vector $vl$. Then, S\ref{sort_and_name_method:3} uses a sorter to merge these partial results, where the sorter maintains a minimal heap in RAM to determine the current smallest substring in $vs_1, vs_2, ..., vs_k$ and $vl$. For any two substrings in the heap, their lexicographical order can be determined in $\mathcal{O}(D)$ time by literally comparing their characters from left to right. This sorting method achieves a good performance when most of the S*-type substrings are short. This makes it feasible in practice because the average length of S*-type substrings in real-world datasets are typically small.

Next, we describe how to name S*-type substrings at the same time when they are being sorted. Given two S*-type substrings successively popped from the minimal heap, we make a literal comparison of them to check their equality if they are both short; otherwise, we use the naming method of SAIS-PQ to determine their equality in S\ref{sort_and_name_method:2}. Specifically,
substrings are sorted by comparing their heading characters and the names of their sorted successors in S\ref{sort_and_name_method:2}, where two substrings are of the same name if and only if their heading characters and the names of their successors are both equal. By recording the names of long S*-type substrings in their sorted order, it takes constant time to check equality of two lexicographically neighboring long substrings in S\ref{sort_and_name_method:3}.

\subsection{Discussion} \label{subsec:further_discussion}

Section~\ref{sec:experiments} shows that, by using the new substring sorting and naming methods, our program for the enhanced  DSA-IS has an advantage over that for eSAIS in terms of space and I/O volume, but its average peak disk use is still higher than that for fSAIS by around $5n$. This is mainly due to the use of external-memory containers (vectors, sorters and priority queues) provided by the STXXL library~\cite{Dementiev2007}. More specifically, the STXXL's containers fail to free the disk space for saving temporary data immediately after they are no longer to use. This implementation issue can be solved by storing temporary data into multiple files and deleting each file when it is obsolete. In this way, our program can be further improved to achieve a space performance similar to that for fSAIS.

@@

\subsection{Prior Art} \label{sec:checkers:prior_art}
The widespread software packages for DC3 and eSAIS provide users a checker to perform verification. Theorem~\ref{theorem:1} gives the main idea behind this checker.

\begin{theorem} \label{theorem:1}
	$sa[0, n)$ is the SA for $x[0, n)$ if and only if the following conditions are satisfied:\\
	(1) $sa$ is a permutation of $[0, n)$. \\	
	(2) ${\sf r_i < r_j} \Leftrightarrow (x[i], r_{i + 1}) < (x[i], r_{j + 1})$ for $ i, j \in [0, n)$ and $i\ne j$, where $r_i$ and $r_j$ indicates the relative order of ${\sf suf}(i)$ and ${\sf suf}(j)$ among all the suffixes, respectively. \\
\end{theorem}

\begin{IEEEproof}
	The proof is available in~\cite{Dementiev2008a}.
\end{IEEEproof}

This checker performs two passes of integer sorts and each sort involves $n$ fixed-size tuples. When implemented in external memory, it requires large disk space and consumes high I/O volume, resulting in a non-negligible checking overhead. As will be seen in Section~\ref{sec:experiments}, the peak disk use for checking an SA encoded by 40-bit integers is around $21n$, while the I/O volume is no less than $50n$. In this section, we describe two lightweight SA checking methods based on the IS principle. Both of them can be seamlessly integrated into any IS suffix sorting algorithm to build and check SA simultaneously.

@@

\subsubsection{Preliminaries} \label{sec:checkers:method_a:preliminaries}

{\em LCP array.} The LCP array for $sa$, denoted by $lcp$, satisfies that $lcp[0] = 0$ and $lcp[i] = \ell$ for $i \in [1, n)$, where $\ell$ is the length of the longest common prefix of ${\sf suf}(sa[i])$ and ${\sf suf}(sa[i - 1])$.

{\em Suffix and LCP buckets.} All the suffixes in $sa$ are naturally partitioned into multiple buckets. Each bucket contains suffixes starting with a same character and occupies a contiguous interval of $sa$. For example, ${\sf sa\_bkt}(c_0)$ gathers all the suffixes starting with $c_0$. This bucket can be further partitioned into two parts, where the left and right parts only contain the L-type and S-type suffixes, respectively. For short, we denote the left and right parts by ${\sf sa\_bkt_L}(c_0)$ and ${\sf sa\_bkt_S}(c_0)$, respectively. Accordingly, $lcp$ can be also partitioned into multiple buckets and ${\sf lcp\_bkt}(c_0)$ records the LCP-values of suffixes in ${\sf sa\_bkt}(c_0)$. Similarly, we define ${\sf lcp\_bkt_L}(c_0)$ and ${\sf lcp\_bkt_S}(c_0)$ for ${\sf sa\_bkt_L}(c_0)$ and ${\sf sa\_bkt_S}(c_0)$, respectively.

{\em Suffix and LCP arrays for S*-type suffixes.} Let $n_1 = \|x_1\|$, we use $sa^*[0, n_1)$ and $lcp^*[0, n_1)$ to denote the suffix and LCP arrays for S*-type suffixes, respectively. The $i$-th element in $sa^*$ records the starting position of the $(i + 1)$-th smallest S*-type substring, while the $i$-th element in $lcp^*[i]$ records the LCP-value of ${\sf suf}(sa^*[i])$ and ${\sf suf}(sa^*[i - 1])$.

\subsubsection{Idea} \label{sec:checkers:method_a:idea}

During the induction phase, SA-IS first computes $sa^*$ by sorting the starting positions of all the S*-type suffixes with their ranks indicated by $sa_1$ and then induces $sa$ from $sa^*$ by S\ref{induction_phase:1}'-S\ref{induction_phase:3}'.

\begin{enumerate}[S1']

	\item Scan $sa^*$ rightward with $i$ decreasing from $n_1 - 1$ to $0$. For each scanned item $sa^*[i]$, insert it into the rightmost empty position of ${\sf sa\_bkt_S}(x[sa^*[i]])$. \label{induction_phase:1}

@@

	\item Scan $sa$ leftward with $i$ decreasing from $n - 1$ to $0$. For each scanned non-empty item $sa[i]$, insert $sa[i] - 1$ into the rightmost empty position of ${\sf sa\_bkt_S}(x[sa[i] - 1])$ if $t[sa[i] - 1] = 1$.\label{induction_phase:3}

@@

An example of the above inducing process is given in Fig.~\ref{fig:example1}. In line 6, we find the end of each bucket and insert the sorted S*-type suffixes into the corresponding buckets. Then, we find the head of each bucket (marked by the symbol $\wedge$) and scan $sa$ rightward for inducing the order of L-type suffixes. In lines 8-9, we check $\sf suf(14)$ (marked by the symbol $@$) and find its predecessor ${\sf suf(13)}$ is L-type, thus we put ${\sf suf(13)}$ into the current leftmost empty position in ${\sf sa\_bkt_L}(i)$. To step through $sa$ in this way, we get all the L-type suffixes sorted in line 17. After that, we find the end of each bucket and scan $sa$ leftward for inducing the order of S-type suffixes in lines 18-29. When scanning $sa[14]$, we see $x[3]$ is S-type and thus put ${\sf suf(3)}$ into the current rightmost empty position in ${\sf sa\_bkt_S}(i)$. Following the same, we get all the S-type suffixes sorted in line 29.

@@

As stated below, S\ref{induction_phase:1}'-S\ref{induction_phase:3}' also constitute the sufficient and necessary conditions for a correct $sa$.

\begin{lemma} \label{lemma:3}
	For any IS suffix sorting algorithm, its output $sa[0, n)$ is the SA for the input string $x[0, n)$ if and only if the following conditions are satisfied: \\ 
	(1) $sa^*$ is correct. \\
	(2) $sa[i]$ is equal to the value calculated by SS\ref{induction_phase:1}-SS\ref{induction_phase:3}. \\

@@

Suppose $sa^*$ is correct\footnote{The correctness of $sa^*$ can be checked by a sparse SA checker, such as~\cite{wu2017}. The checking processes for $sa^*$ and $sa$ can be executed in parallel.}, this lemma suggests a method to check $sa$ when it is induced from $sa^*$ during the induction phase. Specifically, in S\ref{induction_phase:2}' and S\ref{induction_phase:3}', each item induced into $sa$ will be scanned later to induce the order of its predecessor. If the induced and scanned values for each item are equal, then $sa$ is correct according to Lemma~\ref{lemma:3}. The problem to be solved here is that when an item is induced into a bucket, its value in $sa$ will not be scanned at once. Our solution for this is to integrate the building and checking processes into a whole, by ensuring the sequence of items induced into a bucket is identical to that scanned later. For the purpose, we increasingly compute the fingerprints of both sequences and check their equality at the end of the induction phase. If the two fingerprints for each bucket are equal, then the second condition of Lemma~\ref{lemma:3} will be seen with a high probability. As a result, $sa$ can be built and probabilistically checked at the same time, for this we have Corollary~\ref{corollary:1}.

@@

@@

We introduce Lemma~\ref{lemma:1} before the presentation of Method B. This statement has been applied to deriving Theorem~\ref{theorem:1} for comparing two suffixes given the order of their successors already known.

Below we describe the idea of Method B and prove its correctness using Lemma~\ref{lemma:1} and Theorem~\ref{theorem:1}, where $\overline{sa^*}$ is retrieved from $sa$ and it records the starting positions of all the S*-type suffixes in their sorted order.

>>>>

Suffix array~(SA) can be built by the induced sorting~(IS) method on both internal and external memory models. We propose two methods that enable any IS suffix sorter to build and check an SA simultaneously. The first method can check both suffix and longest common prefix~(LCP) arrays for strings drawn from constant alphabets, while the second method is applicable to checking SA for strings of any alphabets. By combining our methods with the Karp-Rabin fingerprinting function, we design two probablistic algorithms that perform verification with a neligible error probability. These algorithms are rather lightweight in terms of time and space. Particularly, the algorithm designed by the second method only takes linear time and constant space when running in external memory. For a further study, we integrate the checking algorithm designed by the second method into the existing suffix sorter DSA-IS to evaluate the checking overhead, where the time, space and I/O consumptions for verification are considerably smaller than that for construction in the experiments. This convinces us that the proposed checking methods, together with the state-of-the-art IS suffix sorters, could constitute efficient solutions for both construction and verification.

@@

A suffix array can be built in linear time and space by the internal-memory algorithm SA-IS~\cite{Nong11}. According to the IS principle, the relative order of two unsorted suffixes is determined by sequentially comparing the heading characters and the sorted succeeding suffixes starting with their second characters. Recently, the IS method has been applied to designing three disk-based suffix sorting algorithms eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15}, which have a better time complexity than the other alternatives~(e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014} and pSAscan~\cite{Karkkainen2015}) but suffer from a bottleneck due to the large disk space for obtaining the heading characters of unsorted suffixes and the ranks of their successors in a disk-friendly way. More specifically, the average peak disk use to construct an SA encoded by 40-bit integers for pSAscan is $7.5n$, while that for eSAIS, DSA-IS and SAIS-PQ are $24n$, $20n$ and $15n$, respectively. The poor space performance is mainly because that the current programs for these algorithms fail to free the disk space for temporary data even when the data is no longer to use. A dramatic improvement can be achieved by splitting a file into multiple pieces and deleting each piece immediately when the temporary data in it is not needed any more. This technique has been used to implement a new IS suffix sorting algorithm fSAIS~\cite{Karkkainen2017}. As reported, the engineering of fSAIS consumes no more than $8n$ disk space for constructing an SA encoded by 40-bit integers, indicating a great potential for optimizing our programs for DSA-IS and SAIS-PQ.

A constructed SA should be checked to avoid computation errors caused by implementation bugs and/or hardware malfunctions. Currently, the software packages for DC3 and eSAIS provide users a checker designed by the method proposed in~\cite{Dementiev2008a}. The overhead for this checker is rather high because it takes two external-memory sorts for arranging $\mathcal{O}(n)$ fixed-size tuples according to their integer keys. Against the background, we propose in this paper two methods that enable any IS suffix sorter to check an SA when it is being built. The first method can check both suffix and LCP arrays for strings drawn from constant alphabets, while the second method can check SA for strings drawn from any alphabets. We augment these methods with the Karp-Rabin fingerprinting function~\cite{Karp1987} to design two probabilistic algorithms, in terms of that their checking results are wrong with a negligible probability. These algorithms are rather lightweight in terms of time and space. Particularly, the algorithm designed by the second method only takes linear time and constant space to run using external memory. For a further study, we integrate this algorithm into DSA-IS to evaluate its checking overhead, where the design of DSA-IS has been adapted by new substring sorting and naming methods for a better performance. As demonstrated in our experiments, the time, space and I/O consumptions for verification by this algorithm are negligible to that for construction by the adapted DSA-IS. We are convinced that the proposed SA checkers, together with the state-of-the-art IS suffix sorters, could constitute efficient solutions for both construction and verification.

The rest of this paper is organized as follows. Section~\ref{sec:preliminaries} introduces some basic notations and symbols. Section~\ref{sec:builder} gives an overview of IS suffix sorting algorithms and the details of new substring sorting and naming methods. Section~\ref{sec:checkers} describes the ideas of checking methods and algorithms. Sections~\ref{sec:experiments} and~\ref{sec:conclusion} show the experimental results and the concluding remarks, respectively.

\section{Preliminaries}~\label{sec:preliminaries}

Given a string $x[0,n)$ drawn from a full-ordered alphabet $\Sigma$, we assume the ending character $x[n - 1]$ to be unique and lexicographically smaller than any other characters in $x$. For convenience, we denote by ${\sf suf}(i)$ and ${\sf sub}(i, j)$ the suffix running from $x[i]$ to $x[n-  1]$ and the substring running from $x[i]$ to $x[j]$, respectively. The following notations are also used in our presentation.

Characters in $x$ are classified into two categories: L-type and S-type. We say $x[i]$ is S-type if (1) $i = n - 1$ or (2) $x[i] = x[i + 1]$ and $x[i + 1]$ is S-type; otherwise $x[i]$ is L-type. Further, if $x[i]$ and $x[i - 1]$ are respectively S-type/L-type and L-type/S-type, then $x[i]$ is also called S*-type/L*-type. We use an array $t$ to record the type of all the characters in $x$, where $t[i] = 1$ or $0$ if $x[i]$ is S-type or L-type, respectively. A substring or suffix is of the same type as its heading character. In this paper, we only consider substrings ending with an S*-type character.

Given two characters $x[i]$ and $x[i + 1]$, we say $x[i]$ is the predecessor of $x[i + 1]$ and $x[i + 1]$ is the successor of $x[i]$. Accordingly, we define the predecessor-successor relationship between ${\sf suf}(i)/{\sf sub}(i, j)$ and ${\sf suf}(i + 1)/{\sf sub}(i + 1, j)$.

We partition $x$ into S*-type substrings and replace each substring with its name to produce the reduced string $x_1$, where the name indicates the rank of the corresponding S*-type substring among all.

The suffix array $sa$ indicates the lexical order of all the suffixes in $x$, where $sa[i]$ records the starting position of the $(i + 1)$-th smallest suffix. We also define the suffix array $sa_1$ for $x_1$ following the same way.

The LCP array $lcp$ indicates the LCP-values of all the neighboring suffixes in $sa$, where $lcp[0] = 0$ and $lcp[i] = \ell$ for $i \in [1, n)$, where $\ell$ is the length of longest common prefix of ${\sf suf}(sa[i])$ and ${\sf suf}(sa[i - 1])$.

All the suffixes in $sa$ are naturally partitioned into multiple buckets. Each bucket occupies a contiguous interval in $sa$ and contains all the suffixes starting with a certain character. Without loss of generality, we denote by ${\sf sa\_bkt}(c_0)$ the bucket for suffixes starting with $c_0$. Notice that ${\sf sa\_bkt}(c_0)$ can be further divided into two parts, where the left part ${\sf sa\_bkt_L}(c_0)$ and the right part ${\sf sa\_bkt_S}(c_0)$ contain the L-type and S-type suffixes, respectively. We also partition $lcp$ into buckets such that ${\sf lcp\_bkt}(c_0)$, ${\sf lcp\_bkt_L}(c_0)$ and ${\sf lcp\_bkt_S}(c_0)$ record the LCP-values of suffixes in ${\sf sa\_bkt}(c_0)$, ${\sf sa\_bkt_L}(c_0)$ and ${\sf sa\_bkt_S}(c_0)$, respectively.

Let $n_1 = \|x_1\|$, we use $sa^*[0, n_1)$ and $lcp^*[0, n_1)$ to denote the suffix and LCP arrays for S*-type suffixes in $x$, respectively. Specifically, $sa^*[i]$ records the starting position of the $(i + 1)$-th smallest S*-type substring, while $lcp^*[i]$ records the LCP-value of ${\sf suf}(sa^*[i])$ and ${\sf suf}(sa^*[i - 1])$.

@@

@@

	\caption{The overview of IS suffix sorting algorithms	.}

@@

As shown in Algorithm~\ref{alg:1}, an IS suffix sorting algorithm first performs a reduction phase to produce the reduced string $x_1$. If there exist two equal characters in $x_1$, then it recursively calls the reduction phase with $x_1$ as input; otherwise, all the S*-type suffixes in $x$ are already sorted and it performs an induction phase to produce the final $sa$. During the execution of two phases, the order of substrings/suffixes is determined by comparing the heading characters and the ranks of their successors according to the IS principle. This involves frequent random accesses to $x$ and $sa$ for retrieving the heading characters of unsorted substrings/suffixes and the ranks of their sorted successors. It can be done very quickly if both $x$ and $sa$ are wholly loaded into RAM; otherwise, each disk access takes an individual I/O operation, leading to a severe performance degradation. This problem is solved in DSA-IS by conducting two steps:

\begin{enumerate}[S1]
	\item Split $x$ into blocks and sort substrings/suffixes of each block by calling SA-IS, the heading characters in use are copied to external memory in their access order. \label{dsais_sorting_method:1}
	\item Sort the substrings/suffixes of $x$ by their heading characters and the ranks of their successors. Organize the sorted substrings/suffixes in an external-memory heap and induce their predecessors into the heap according to the IS principle, where the heading characters of the induced substrings/suffixes are retrieved from external memory by sequential I/O operations.\label{dsais_sorting_method:2}
\end{enumerate}

As shown in Section~\ref{sec:experiments}, our program for DSA-IS requires less disk space than that for eSAIS, but the former runs slower than the latter caused by the large I/O volume for sorting and naming S*-type substrings during the reduction phase. For this, we present new substring sorting and naming methods in the next subsection.

@@

All the S*-type substrings are classified into long and short categories with respect to whether or not containing more than $D$ characters. The new substring sorting method mainly consists of the three steps below:

\begin{enumerate}[S1']
	\item Sort the long of each block by S\ref{dsais_sorting_method:1}. During the process, copy the short to external memory in their sorted order.~\label{dsaism_sorting_method:1}

	\item Sort the long of $x$ by S\ref{dsais_sorting_method:2}. During the process, copy the leftmost $D$ characters of the long to external memory in their sorted order.~\label{dsaism_sorting_method:2}

	\item Merge the short and long by a multi-way sorter.~\label{dsaism_sorting_method:3}
\end{enumerate}

After the first two steps, the short S*-type substrings of each block and the long S*-type substrings of $x$ are sorted and separately organized as a sequence in external memory. Assume $x$ is split into $k$ blocks, the multi-way sorter in S\ref{dsaism_sorting_method:3}' maintains an internal-memory heap to cache the current smallest of each sequence and continually retrieve the top item from the heap to determine the lexical order of substrings from different sequences. The heap contains at most $k + 1$ substrings at any point of time and it compares any two substrings in $\mathcal{O}(D)$ time by a literal string comparison\footnote{If the leftmost $D$ characters of a long S*-type substring is equal to a short S*-type substring, then the short is lexicographically greater than the long.}. Theoretically, the sorting method has a good performance if the majority of S*-type substrings in $x$ are short with a small $D$. This is commonly satisfied in practice, because the average length of S*-type substrings is typically small in real-world datasets.

In what follows, we describe a method for naming the S*-type substrings during the above sorting process. The key point here is to check equality of two substrings successively popped from the heap in S\ref{dsaism_sorting_method:3}' as fast as possible. If either of the two substrings is short, our solution is to literally compare them in $\mathcal{O}(D)$ time. Otherwise, if both of them are long, then we assign names to all the long S*-type substrings using the technique proposed for SAIS-PQ in S\ref{dsaism_sorting_method:2}' and compare two successively popped long S*-type substrings by their names. This technique is based on the idea that two substrings are of the same name if their heading characters and the names of their successors are both equal\footnote{We refer interested readers to~\cite{Liu15} for more information.}.

The experimental results in Section~\ref{sec:experiments} indicate that, by using the new substring sorting and naming methods, the adapted DSA-IS, called DSA-IS+, only takes half as much disk space as eSAIS. Also, the peak disk use of the program for DSA-IS+ can be further reduced to a level comparable to that of the program for fSAIS.

@@

\subsection{Prior Art} \label{sec:checkers:prior_art}

The software packages for DC3 and eSAIS provide users a checker to ensure the correctness of their outputs. The following statements describe the idea behind this checker.

\begin{theorem} \label{theorem:1}
	$sa[0, n)$ is the SA for $x[0, n)$ if and only if the following conditions are satisfied:\\
	(1) $sa$ is a permutation of $[0, n)$. \\	
	(2) ${\sf r_i < r_j} \Leftrightarrow (x[i], r_{i + 1}) < (x[i], r_{j + 1})$ for $ i, j \in [0, n)$ and $i\ne j$, where $r_i$ and $r_j$ represent the ranks of ${\sf suf}(i)$ and ${\sf suf}(j)$ among all the suffixes, respectively. \\
\end{theorem}

\begin{IEEEproof}
	Theorem~\ref{theorem:1} is derived from Lemma~\ref{lemma:1}. The proof is available in~\cite{Dementiev2008a}.
\end{IEEEproof}

This checker performs two passes of integer sorts and each sort involves $n$ fixed-size tuples. When implemented in external memory, it consumes large disk space and I/O volume, resulting in a high checking overhead. For example, we can observe from Section~\ref{sec:experiments} that, the peak disk use and the I/O volume for checking an SA encoded by 40-bit integers are around $26n$ and $53n$, respectively. In this section,  we describe two SA checking methods based on the IS principle, both of them can be seamlessly integrated into any IS suffix sorting algorithm to build and check an SA at the same time.

@@

Given that each S*-type substring is unique in $x$, SA-IS first computes $sa^*$ by sorting the starting positions of all the S*-type suffixes with their ranks indicated by $sa_1$ and then induces $sa$ from $sa^*$ by S\ref{induction_phase:1}''-S\ref{induction_phase:3}'' during the inducting phase.

\begin{enumerate}[S1'']

	\item Clear $sa$. Scan $sa^*$ leftward with $i$ decreasing from $n_1 - 1$ to $0$. For each scanned item $sa^*[i]$, insert it into the rightmost empty position of ${\sf sa\_bkt_S}(x[sa^*[i]])$. \label{induction_phase:1}

@@

	\item Clear ${\sf sa\_bkt_S}(c)$ for $c \in \Sigma$. Scan $sa$ leftward with $i$ decreasing from $n - 1$ to $0$. For each scanned non-empty item $sa[i]$, insert $sa[i] - 1$ into the rightmost empty position of ${\sf sa\_bkt_S}(x[sa[i] - 1])$ if $t[sa[i] - 1] = 1$.\label{induction_phase:3}

@@

We illustrate a running example of the above process in Fig.~\ref{fig:example1} and show more details for a better understanding. In line 6, we find the ending position of each bucket and insert the sorted S*-type suffixes into the right part of the corresponding buckets. Afterward, we first find the heading position of each bucket (marked by the symbol $\wedge$) and scan $sa$ rightward for inducing the order of L-type suffixes as following. In lines 8-9, we check $\sf suf(14)$ (marked by the symbol $@$) and find its predecessor ${\sf suf(13)}$ is L-type, thus we put ${\sf suf(13)}$ into the current leftmost empty position in ${\sf sa\_bkt_L}(i)$. To step through $sa$ in this way, we get all the L-type suffixes sorted in line 17. After that, we find the ending position of each bucket and scan $sa$ leftward for inducing the order of S-type suffixes in lines 18-28. When scanning $sa[14]$, we see $x[3]$ is S-type and thus put ${\sf suf(3)}$ into the current rightmost empty position in ${\sf sa\_bkt_S}(i)$. Following the same, we get all the S-type suffixes sorted in line 28.

@@

As stated below, S\ref{induction_phase:1}''-S\ref{induction_phase:3}'' also constitute the sufficient and necessary conditions for a correct $sa$.

\begin{lemma} \label{lemma:3}
	For any IS suffix sorting algorithm, its output $sa[0, n)$ is the SA for the input string $x[0, n)$ if and only if the following conditions are satisfied: \\ 
	(1) $sa^*$ is correct. \\
	(2) $sa[i]$ is equal to the value calculated by S\ref{induction_phase:1}''-S\ref{induction_phase:3}''. \\

@@

Suppose $sa^*$ is correct\footnote{The correctness of $sa^*$ can be checked by a sparse SA checker, such as~\cite{wu2017}. It should be noticed that the checking processes for $sa^*$ and $sa$ can be executed in parallel.}, this lemma suggests a method to check $sa$ when it is induced from $sa^*$ during the induction phase. Specifically, in S\ref{induction_phase:2}'' and S\ref{induction_phase:3}'', each item induced into $sa$ will be scanned later to induce the order of its predecessor. If the induced and scanned values for each item are equal, then $sa$ is correct according to Lemma~\ref{lemma:3}. The problem to be solved here is that when an item is induced into a bucket, its value in $sa$ will not be scanned at once. One solution for this is to record the induced and scanned values separately during the building process and check their equality afterward. Our solution is to integrate the building and checking processes into a whole, by ensuring the sequence of items induced into a bucket is identical to that scanned later. For the purpose, we can use a fingerprinting function to increasingly compute the fingerprints of both sequences and check their equality in constant time at the end of induction phase. If the two fingerprints for each bucket are equal, then the second condition of Lemma~\ref{lemma:3} will be seen with a high probability. As a result, $sa$ can be built and probabilistically checked at the same time. Following the previous discussion, we design a probabilistic algorithm based on  Corollary~\ref{corollary:1}.

@@

We point out that Method A can be also applied to checking $lcp$ when it is being built from $lcp^*$ following the IS principle~\cite{Fischer11}. In general, this is done by comparing the sequences of induced and scanned items for all the buckets in $lcp$.

@@

{\color{red} 1. $\overline{sa^*}$ is a subset of the final $sa$, what is $sa^*[i]$? These two symbols have not been defined clearly and are confusing.

2. Notice: The whole algorithm for induced sorting suffixes consists of many processes, the inducing process is only one of them. The checking takes place only at the top level by reusing the inducing process there. For the checking algorithm to work, it must be correctly implemented, this implies that the inducing process at the top level is also implicitly assumed to be correctly implemented. This needs to be specified/assumed for clarity.
}

{ \color{blue} The definition of $sa^*$ was introduced in Section~\ref{sec:preliminaries}. An example was given on the left for better understanding.}
